Meta analysis of differential expression results from AMP-AD (Control vs AD)

### 2016-06-06 code/Rmd/MetaAnalysis.Rmd
minor fix

knit2synapse::knitToFolderEntity(file = "./MetaAnalysis.Rmd", 
                                 parentId ="syn5569102",
                                 entityName = 'MetaAnalysis of Control vs AD Data')
```

@@

# Load required libraries
library(synapseClient)
library(RColorBrewer)
library(ggplot2)
library(data.table)
library(plyr)
library(dplyr)
library(psych)
library(tidyr)
library(Matrix)

library(limma)
library(sva)
library(RankProd)

library(knitr)
library(rGithubClient) ## Needs the dev branch

## Requires ggdendro
devtools::source_url("https://raw.githubusercontent.com/chr1swallace/random-functions/master/R/ggplot-heatmap.R")

@@

# source utility files from ../R/lib folder
source('../R/lib/rownameToFirstColumn.R')
# file.sources = list.files('../R/lib/rownfile.sources',pattern="row*.R", full.names = T)
# tmp = sapply(file.sources,source,.GlobalEnv)

@@

```{r fxns}
# Set ggplot theme
my.theme = theme(title = element_text(size = 24), 
                 axis.text = element_text(size = 20), 
                 axis.text.x = element_text(angle = 90, hjust=1, vjust = 0.5))

# Utility function to download files from synapse
download.file <- function(id){ tmp = fread(synGet(as.character(id))@filePath, data.table=F, header=T);  }

# Utility function for combining p-values using Fishers inverse chi-square method
fishersPvalue <- function(pvalue){
  pvalue = -2*log(pvalue, base = exp(1))
  S <- 0
  N <- dim(pvalue)[2]

@@

### Download differential expression data
```{r download.data}
# Get all differential expression files and their synIds
DEXP_ID = c(ROSMAP.PFC = 'syn5608845',
            MSSB = 'syn5609009',
            MAYO.CER = 'syn5609850',
            MAYO.TC = 'syn5609813',
            MAYO.CER.AR = 'syn5607457',
            MAYO.TC.AR = 'syn5607475')

# Download differential expression data
DEXP = lapply(DEXP_ID, download.file)

# Extract subset of differential expression data
DEXP$ROSMAP.PFC = DEXP$ROSMAP.PFC %>% 
  dplyr::filter(Comparison == "cogdx1-cogdx4") %>%
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MSSB.FP = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_10.Control-BM_10.AD") %>%
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MSSB.STG = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_22.Control-BM_22.AD") %>%
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MSSB.PHG = DEXP$MSSB %>% 
  dplyr::filter(Comparison == "BM_36.Control-BM_36.AD") %>%
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MSSB = NULL

DEXP$MAYO.CER = DEXP$MAYO.CER %>% 
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MAYO.TC = DEXP$MAYO.TC %>% 
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MAYO.CER.AR = DEXP$MAYO.CER.AR %>% 
  dplyr::select(hgnc_symbol, P.Value, logFC)

DEXP$MAYO.TC.AR = DEXP$MAYO.TC.AR %>% 
  dplyr::select(hgnc_symbol, P.Value, logFC)
```

@@

### Fisher-Brown's meta analysis
```{r fisher.brown}
Pfb = data.frame(hgnc_symbols = Pval[,1], 
                 Pfb = p.adjust(fishersPvalue(Pval[,-(1)]), method = 'fdr'))

@@

# Store pvalues
tmp = join_all(list(Pmin,Pfb))
write.table(tmp, file = './Pval.tsv', sep = '\t', row.names=F, quote=F)
FILE_OBJ = File("./Pval.tsv", name = 'Meta Differential Expression Analysis FDR', parentId = CODE$properties$id)
FILE_OBJ = synStore(FILE_OBJ, 
                    used = as.character(DEXP_ID),
                    executed = thisFile,
                    activityName = 'Differential expression analysis of control-AD')

# Store differentially expressed genes
write.table(rownameToFirstColumn(Adj.P.Value, 'hgnc_symbol'), file = './DiffGenes.tsv', sep = '\n', row.names=F, quote=F, col.names = F)
FILE_OBJ = File("./DiffGenes.tsv", name = 'Differentially expressed genes FDR (individual studies)', parentId = CODE$properties$id)
FILE_OBJ = synStore(FILE_OBJ, 
                    used = as.character(DEXP_ID),
                    executed = thisFile,
                    activityName = 'Differential expression analysis of control-AD')
```

>>>>

knit2synapse::knitToFolderEntity(file = "./MetaAnalysis.Rmd", 
                                 parentId ="syn5569102",
                                 entityName = 'MetaAnalysis of Control vs AD from all three brain banks')
```

@@

# Load required libraries
library(CovariateAnalysis)
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)

library(RColorBrewer)
library(ggplot2)
library(ComplexHeatmap)

library(RankProd)
library(githubr)

@@

@@

```{r fxns}
# Utility function for combining p-values using Fishers inverse chi-square method
fishersPvalue <- function(pvalue){
  pvalue = -2*log(pvalue, base = exp(1))
  S <- 0
  N <- dim(pvalue)[2]

@@

### Get covariates/sample metadata from synapse
```{r covariates}
covariates.id = c('ROSMAP' = 'syn6114444',
                  'MSBB' = 'syn6131421',
                  'MAYO' = 'syn6128267')

covariates = lapply(covariates.id, downloadFile)

covariates$ROSMAP = covariates$ROSMAP %>%
  dplyr::select(Sampleid, msex, cogdx) %>%
  dplyr::rename(ID = Sampleid, Gender = msex, Status = cogdx) %>%
  dplyr::mutate(Gender = factor(Gender, labels = c('1' = 'MALE','2'='FEMALE')),
                Status = factor(Status, labels = c('1' = 'NCI', '2' = 'MCI', '3' = 'MCI.CI', 
                                                   '4' = 'AD', '5' = 'AD.CI', '6' = 'OD')),
                BrainRegion = 'DLPFC') %>%
  dplyr::filter(Status %in% c('NCI', 'AD'))

covariates$MSBB = covariates$MSBB %>%
  dplyr::select(SampleId, SEX, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
  dplyr::rename(ID = SampleId, Gender = SEX, Status = Dx) %>%
  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
                Status = factor(Status),
                BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
  dplyr::filter(Status %in% c('ND', 'SD'))

covariates$MAYO = covariates$MAYO %>%
  dplyr::select(ID, Gender, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx'), sep = '\\.') %>%
  dplyr::rename(Status = Dx) %>%
  dplyr::mutate(Gender = factor(Gender, labels = c('M' = 'MALE','F'='FEMALE')),
                Status = factor(Status)) %>%
  dplyr::filter(Status %in% c('Control', 'AD'))

covariates = rbindlist(covariates, use.names = T, fill = T, idcol = 'Study')

```

### Get logCPM counts from synapse
```{r counts}
logcpm.id = c('ROSMAP' = 'syn6114447' ,
              'MSBB' = 'syn6131419',
              'MAYO' = 'syn6128269')

logcpm = lapply(logcpm.id, downloadFile) %>%
  lapply(function(x){
    x = dplyr::select(x, -hgnc_symbol)
    }) %>%
  join_all(type = 'full') %>%
  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
logcpm[is.na(logcpm)] = dplyr::select(logcpm, -Gene.ID, -ensembl_gene_id) %>%
  min
```

### Download differential expression data
```{r download.data}
diffexp.id = c('ROSMAP' = 'syn6114453',
               'MSBB' = 'syn6131423',
               'MAYO' = 'syn6128273')

diffexp = lapply(diffexp.id, downloadFile)

diffexp$ROSMAP = diffexp$ROSMAP %>%
  filter(Comparison == 'cogdx1-cogdx4') %>%
  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val) %>%
  dplyr::mutate(Tissue = 'DLPFC', Contrast = 'NCI-AD')

diffexp$MSBB = diffexp$MSBB %>%
  dplyr::filter(Name == 'ND-SD') %>%
  dplyr::select(Gene.ID, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion, labels = c('BM_10' = 'FP', 'BM_22' = 'STG', 
                                                             'BM_36' = 'PHG', 'BM_44' = 'IFG'))) %>%
  dplyr::rename(Tissue = BrainRegion, Contrast = Name)

diffexp$MAYO = diffexp$MAYO %>%
  dplyr::filter(Name == 'Control-AD') %>%
  dplyr::select(genes, logFC, P.Value, adj.P.Val, BrainRegion, Name) %>%
  dplyr::rename(Tissue = BrainRegion, Gene.ID = genes, Contrast = Name)

diffexp = rbindlist(diffexp, use.names = T, fill = T, idcol = 'Study') %>%
  tidyr::separate(Gene.ID, c('ensembl_gene_id','position'),sep = '\\.')

diffexp$DE = 'NONE'
diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC >= log2(1.2)] = 'UP'
diffexp$DE[diffexp$adj.P.Val <= 0.05 & diffexp$logFC <= -log2(1.2)] = 'DOWN'

pl = list()
for (study in c('ROSMAP', 'MSBB', 'MAYO')){
  p = filter(diffexp, Study == study) %>%
    ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = DE)) 
  p = p + scale_colour_manual(values = c("green","grey", "red"))
  p = p + geom_point() + theme(legend.position = 'None')
  p = p + facet_grid(Study~.+Tissue)
  pl[[study]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Get hgnc_symbol mapping from biomaRt
```{r biomart}
# Define biomart object
mart <- useMart('ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl", host = 'dec2015.archive.ensembl.org')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", 
                       values = unique(diffexp$ensembl_gene_id),
                       mart = mart) %>%
  data.table()

# Add hgnc_symbols to the logcpm matrix
logcpm = logcpm %>%
  left_join(Ensemble2HGNC)

# Add hgnc_symbols to the diffexp matrix
diffexp = diffexp %>%
  left_join(Ensemble2HGNC)
```

### Pathway enrichment
```{r pathway.enrichment}
# Differentially expressed in ROSMAP
## Up regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'DLPFC') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x3h

## Down regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC<0, Tissue == 'DLPFC') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6x4e

# Differentially expressed in MSSM FP
## Up regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC >0, Tissue == 'FP') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xcm

## Down regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'FP') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')

# Differentially expressed in MSSM STG
## Up regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'STG') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')

## Down regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'STG') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')

# Differentially expressed in MSSM PHG
## Up regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'PHG') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xco

## Down regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge

# Differentially expressed in MAYO CER
## Up regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC > 0, Tissue == 'CER') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here:

## Down regulated
tmp = filter(diffexp, adj.P.Val <= 0.05, logFC < 0, Tissue == 'PHG') %>%
  dplyr::left_join(Ensemble2HGNC %>% as.data.table()) %>%
  filter(hgnc_symbol != '')
### Enrichr results are here: http://amp.pharm.mssm.edu/Enrichr/enrich?dataset=6xge
```

### Meta Analysis
```{r metaanalysis}
P.Value = diffexp %>% 
  # filter(Tissue %in% c('DLPFC', 'PHG', 'TCX')) %>%
  dplyr::select(ensembl_gene_id, P.Value, Tissue, Contrast) %>%
  tidyr::unite(Tissue.Contrast, Tissue, Contrast, sep = '_') %>%
  tidyr::spread(Tissue.Contrast, P.Value)
P.Value[is.na(P.Value)] = 1

min.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
  data.matrix %>% rowMin %>%
  p.adjust(method = 'BH') %>%
  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
             min.pval = .)

max.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
  data.matrix %>% rowMax %>%
  p.adjust(method = 'BH') %>%
  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
             max.pval = .)

fb.pval = dplyr::select(P.Value, -ensembl_gene_id) %>%
  data.matrix %>% fishersPvalue %>%
  p.adjust(method = 'BH') %>%
  data.frame(ensembl_gene_id = P.Value$ensembl_gene_id,
             fb.pval = .)

metaAnal = join_all(list(min.pval, max.pval, fb.pval), type = 'full')
tmp = metaAnal %>% gather(method, pval, -ensembl_gene_id) %>% group_by(method) %>% filter(pval <= 0.05, method == 'fb.pval') %>%
  left_join(Ensemble2HGNC) %>%
  dplyr::select(hgnc_symbol) %>%
  unique
```

@@

```{r fisher.brown}
Pfb = data.frame(hgnc_symbols = Pval[,1], 
                 Pfb = p.adjust(fishersPvalue(Pval[,-(1)]), method = 'fdr'))

@@

# Store covariates in synapse
write.table(covariates, file = 'ALLCovariates.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLCovariates.tsv', 
           name = 'Covariates', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store logcpm in synapse
write.table(logcpm, file = 'ALLLogCPM.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLLogCPM.tsv', 
           name = 'Log CPM (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')

# Store diffexp in synapse
write.table(diffexp, file = 'ALLDiffExp.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('ALLDiffExp.tsv', 
           name = 'Differential Expression (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(c(covariates.id)),
               executed = thisFile,
               activityName = 'MetaAnalysis of AMP-AD reprocessed data')
```
